Comprehensive assessment of the climate-induced water scarcity over continental Chile

### 2023-05-30 article/article.Rmd
escribir abstract

---
title: Comprehensive assessment of the climate-induced water scarcity over continental Chile
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.com
    affiliation: Hemera Centro de Observación de la Tierra,Universidad Mayor
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Hemera Centro de Observación de la Tierra
    organization: Facultad de Ciencias, Universidad Mayor.
    addressline: La Piramide 5750
    city: Santiago
    state: Huechuraba
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes globally. Chile has been affected by persistent water scarcity which has been impacting the hydrological system and vegetation development. Central Chile it is been the focus of research studies due to the diminishing of water supply, nevertheless water deficit is expanded beyond, specifically to the south of the country.

keywords: 
  - keyword1
  - keyword2
journal: "Remote Sensing of Environment"
date: "`r Sys.Date()`"
classoption: preprint, 3p
bibliography: references.bib
biblio-style: "apalike"
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

>>>>

---
title: Comprehensive assessment of the climate-induced water scarcity over continental Chile
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.com
    affiliation: Hemera Centro de Observación de la Tierra,Universidad Mayor
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Hemera Centro de Observación de la Tierra
    organization: Facultad de Ciencias, Universidad Mayor.
    addressline: La Piramide 5750
    city: Santiago
    state: Huechuraba
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes globally. Chile has been affected by persistent water scarcity which is impacting the hydrological system and vegetation development. Central Chile it is been the focus of research studies due to the diminishing water supply, nevertheless water deficit is expanded beyond. We analyze earth observation data for 2000-2023 to do a comprehensive assessment of water scarcity in Chile.  For the analysis, continental Chile was divided into five zones from north to south.  We used the time series of MODIS for land cover change (LULC), fraction of vegetation cover (FVC), and to derive the drought index zcNDVI and zcET (standardized anomaly of cumulative NDVI and ET through the growing season); and CHELSA v2.1 dataset to derive the standardized precipitation evapotranspiration index (SPEI). We evaluate the interconnection between the productivity drought index (zcNDVI) and water demand (zcET), supply (SPEIs) and FVC. Finally, we analyzed the temporal correlation of the drought indices with total water storage (TWS) from GRACE (Gravity Recovery and Climate Experiment). Our results showed that ...

keywords: 
  - keyword1
  - keyword2
journal: "Remote Sensing of Environment"
date: "`r Sys.Date()`"
classoption: preprint, 3p
bibliography: references.bib
biblio-style: "apalike"
linenumbers: false
numbersections: true
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
  word_document: default
---

### 2023-05-30 article/article.tex
escribir abstract

  \begin{abstract}
  Human-induced greenhouse gas emissions have increased the frequency
  and/or intensity of some weather and climate extremes globally. Chile
  has been affected by persistent water scarcity which has been
  impacting the hydrological system and vegetation development. Central
  Chile it is been the focus of research studies due to the diminishing
  of water supply, nevertheless water deficit is expanded beyond,
  specifically to the south of the country.
  \end{abstract}
    \begin{keyword}
    keyword1 \sep 
    keyword2
  \end{keyword}

@@

\caption{\label{tab:unnamed-chunk-1}Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{r>{}r>{}r>{}r>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Trend of change [$km^2 year^{-1}$]} \\
\cmidrule(l{3pt}r{3pt}){2-7}
zone & Shrubland & Savanna & Grassland & Barren land & Forest & Cropland\\
\midrule
norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671073045199.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671013721869.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671079e41f67.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671075ebd9d8.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710524640a1.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671012073e8a.pdf}\\
norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671061580771.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101832bef9.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967105771ba87.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671042ea518f.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710607baab3.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967106f91099.pdf}\\
zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967105c411aeb.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967108a6119a.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671016b408b2.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967103ee6f936.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671074c30c6.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671056667823.pdf}\\
zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967106cc832ad.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671034e3692f.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967104ac90971.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967107f7df47e.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967104e385249.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671013698bfb.pdf}\\
zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710491ed82a.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710190c87cf.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101d6ab929.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671033cd1ca4.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101f22c71d.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710336ead1.pdf}\\
\bottomrule
\end{tabular}}
\end{table}

>>>>

  \begin{abstract}
  Human-induced greenhouse gas emissions have increased the frequency
  and/or intensity of some weather and climate extremes globally. Chile
  has been affected by persistent water scarcity which is impacting the
  hydrological system and vegetation development. Central Chile it is
  been the focus of research studies due to the diminishing water
  supply, nevertheless water deficit is expanded beyond. We analyze
  earth observation data for 2000-2023 to do a comprehensive assessment
  of water scarcity in Chile. For the analysis, continental Chile was
  divided into five zones from north to south. We used the time series
  of MODIS for land cover change (LULC), fraction of vegetation cover
  (FVC), and to derive the drought index zcNDVI and zcET (standardized
  anomaly of cumulative NDVI and ET through the growing season); and
  CHELSA v2.1 dataset to derive the standardized precipitation
  evapotranspiration index (SPEI). We evaluate the interconnection
  between the productivity drought index (zcNDVI) and water demand
  (zcET), supply (SPEIs) and FVC. Finally, we analyzed the temporal
  correlation of the drought indices with total water storage (TWS) from
  GRACE (Gravity Recovery and Climate Experiment). Our results showed
  that \ldots{}
  \end{abstract}
    \begin{keyword}
    keyword1 \sep 
    keyword2
  \end{keyword}

@@

\caption{\label{tab:unnamed-chunk-1}Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{r>{}r>{}r>{}r>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Trend of change [$km^2 year^{-1}$]} \\
\cmidrule(l{3pt}r{3pt}){2-7}
zone & Shrubland & Savanna & Grassland & Barren land & Forest & Cropland\\
\midrule
norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70937146e24b.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709384faf70.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709373919fc5.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709321ad9f19.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709379ea79bd.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093c8ff3c5.pdf}\\
norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709337b6241e.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70937f0ed48a.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70934b4bd377.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70931e4fd155.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932e8b190a.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093305c1274.pdf}\\
zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932bbec995.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093598f66c9.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709370472832.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093ba3805d.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709341ec566.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70933b82960c.pdf}\\
zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70936a87e29b.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70935b745f29.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709363a8f17a.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932c57d58c.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70935fdd829a.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709328c1f26f.pdf}\\
zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709343d0ce35.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093615c29b7.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70933872723d.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709369bca642.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709386fdff0.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70931a654519.pdf}\\
\bottomrule
\end{tabular}}
\end{table}
